|

### 2019-08-06 article/article.Rmd
add text

---
title:  |
        | The Pitfalls of Bottom-Up Accountability: Evidence from Brazil
author: 
- "Danilo Freire[^freire]"
- "Manoel Galdino[^galdino]"
- "Umberto Mignozzetti[^mignozzetti]"
- "Jessica Voigt[^voigt]"
date: \today
thanks: "We thank Guilherme Fasolin, Stephen Herzog and David Skarbek for their helpful comments. We are specially grateful to the staff of Transparência Brasil for their excellent research and technical assistance. Replication data and code are available at <http://github.com/umbertomignozzetti/tdp-accountability>."
abstract: "We study the effect of a mobile phone application designed to increase citizens' ability to monitor public works in Brazil. The app allowed individuals to find delayed school constructions in XXX Brazilian cities and send anonymous messages to politicians requestions information about construction delays and expected completion times. Our results show that the app has a null impact on the completion times subsequently reported by the municipalities. Additionally, we find that few politicians reacted to citizens' requests. These findings suggest it is difficult to motivate bottom-up accountability in new democracies, especially when politicians are unresponsive to non-electoral pressures. This paper has implications for the design of bottom-up accountability mechanisms."
abstractspacing: double
keywords: accountability, Brazil, field experiment, impact evaluation, state capacity, technology
fontsize: 12pt
margin: 2cm
urlcolor: darkblue
linkcolor: Mahogany
citecolor: Mahogany
spacing: double
papersize: a4paper
bibliography: references.bib
biblio-style: apalike
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    number_sections: yes
    keep_tex: no
    toc: no
    toc_depth: 3
    template: article-template.latex
---

@@

[^galdino]: Executive Diretor, Transparência Brazil, São Paulo, SP, Brazil, [mcz.fea@gmail.com](mailto:mcz.fea@gmail.com), <https://www.transparencia.org.br>.

@@

[^voigt]: Data Scientist, Transparência Brasil, São Paulo, SP, Brazil, [voigt.jessica@gmail.com](mailto:voigt.jessica@gmail.com), <https://www.linkedin.com/in/voigtjessica>.

@@

One of the key determinants of efficient public service provision is a robust accountability system [@besley2003incentives; @cameron2004public; @ferejohn1986incumbent; @o1998horizontal]. In its standard definition, accountability is understood as the process of holding authorities responsible for their actions [@finer1941administrative; @mulgan2000accountability; @o1990bureaucratic]. Past studies show that accountability ensures that politicians act on behalf of voters [@freire2010ngp; @moncrieffe1998reconceptualizing], reduces the opportunities for rent-seeking and corruption [@deininger2005does; @wenar2006accountability] and improves the efficiency of the public sector [@adsera2003you; @bjorkman2009power]. Moreover, greater accountability leads to higher economic growth as it limits state discretion in the economy, promotes social cohesion and increases long-term investments in human capital [@benhabib2010economic; @suebvises2018social; @ponzetto2018social].

However, accountability systems can take many forms. One promising model for democratic oversight is _bottom-up monitoring_, in which citizens receive information about the shortcomings of a given project to help them evaluate and pressure underperforming public officials [@kosack2014does; @molina2016community; @raffler2018weakness]. Proponents argue bottom-up accountability is effective because 1) constituents have first-hand information about the outcomes of local policies; 2) citizens have incentives to avoid corruption that directly affects them; 3) policy-makers are particularly sensitive to social punishment from their own communities [@serra2011combining, 570]. In a nutshell, bottom-up accountability offers a potential solution to the principal-agent dilemma in public service by aligning the interests of politicians and bureaucrats with those of the constituency they serve [@barro1973control; @raffler2018weakness, 2].

Here we assess the impact of a technology-driven solution that lowers the costs of evaluating public works and punishing political representatives in Brazil. Developed by Transparência Brasil[^transparencia], the _Tá de Pé_ (TDP) mobile phone application allows Brazilian citizens to know the location of public school construction sites, check their completion status and anonymously request information from competent authorities. TDP users can also take pictures of delayed works and submit them to a group of trained engineers for examination. If the engineers classify the construction as delayed, TDP allows users to send a message to the mayor's office asking for completion estimates and explanations about the construction status. TDP has been online since April 2017 in 1030 Brazilian municipalities and was the winner of the 2016 Google Social Impact grant with more than 200,000 popular votes[^google].

@@

[^google]: About 1,000 Brazilian charities participated in the 2016 Google Social Impact Challenge and an independent selection committee selected 10 of them as finalists. Transparência Brasil was then the highest voted NGO by the public. To know more about the contest, please visit <https://impactchallenge.withgoogle.com/brazil2016> (access: July 2019).

We use the TDP app to conduct **two/three?** experimental interventions. We test whether the app has a positive impact on **school completion rates, school construction status and complaints to competent authorities**. Contrary to our expectations, our experiments have delivered mixed results. **One/Two** of the findings are positive and in line with our theoretical predictions. From August 2017 to February 2019, TDP increased the likelihood of school completion by 6.8 percent (SE = X.X, t-test = XXX). Additionally, we find that pressure over the bureaucracy increased the investments in the schools by 157.5 percent (SE = X.XX, t-test = XXX). In contrast, our results show the app has a null impact on school constructions statistics. Brazilian politicians did not respond to the citizen's requests placed via the application. Therefore, our findings raise questions about the ability of citizens to hold representatives accountable using bottom-up strategies. **Nota**: _Reescrever essa última frase e colocar mais explicações sobre possíveis mecanismos, etc_

@@

Transparência Brasil established a partnership with the Brazilian branch of _Engineers without Borders_ (EWB), a non-governmental organisation dedicated to promote international development[^ewb], to provide unbiased evaluations of school completion rates. Their assessment is based on photos and GPS location sent through the app. All data are collected anonymously. Information on users' alerts are stored on their mobile devices, so citizens can track whether their request has been answered by public authorities.

@@

The TDP has also received feedback from Brazilian computer programmers and policy analists. In 2017 and 2018, Transparência Brasil announced two team programming competitions, called 'Tá de Pé Hackathons', where contributors could fix code bugs and suggest new functionalities to TDP. In total, 75 participants

# Methods
\label{sec:methods}

>>>>

---
title:  |
        | Bottom-Up Accountability and Infrastructure Delivery: Mixed Evidence from Brazil
author: 
- "Danilo Freire[^freire]"
- "Manoel Galdino[^galdino]"
- "Umberto Mignozzetti[^mignozzetti]"
- "Jessica Voigt[^voigt]"
date: \today
thanks: "We thank Fábio Barros, Guilherme Fasolin, Stephen Herzog and David Skarbek for their helpful comments. We are specially grateful to the staff of Transparência Brasil for their excellent research assistance. This research was pre-registered on the EGAP pre-registry tool (<https://egap.org/registration/4505>) and approved by the Brazilian Ministry of Education. The authors thank the 2016 Google Social Impact Challenge for financial support and declare there are no conflict of interest. Replication data and code are available at <http://github.com/umbertomignozzetti/tdp-accountability>."
abstract: "We study the effect of a mobile phone application designed to increase citizens' ability to monitor public works in Brazil. The app allows individuals to find delayed school constructions in 1030 Brazilian cities and send anonymous messages to politicians requesting information about construction delays and completion times. Our results show the app has a null impact on the completion times subsequently reported by the municipalities. Additionally, we find that few politicians react to citizens' requests. These findings suggest it is difficult to motivate bottom-up accountability in new democracies, especially when politicians are unresponsive to non-electoral pressures. This paper has implications for the design of bottom-up accountability mechanisms."
abstractspacing: double
keywords: accountability, Brazil, field experiment, impact evaluation, state capacity, technology
fontsize: 12pt
margin: 
urlcolor: darkblue
linkcolor: Mahogany
citecolor: Mahogany
spacing: double
papersize: a4paper
bibliography: references.bib
biblio-style: apalike
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    number_sections: yes
    keep_tex: no
    toc: no
    toc_depth: 3
    template: article-template.latex
---

@@

[^galdino]: Executive Diretor, Transparência Brazil, São Paulo, SP, Brazil, [mgaldino@transparencia.org.br](mailto:mgaldino@transparencia.org.br), <https://www.transparencia.org.br>.

@@

[^voigt]: Data Scientist, Transparência Brasil, São Paulo, SP, Brazil, [jvoigt@transparencia.org.br](mailto:jvoigt@transparencia.org.br), <https://www.linkedin.com/in/voigtjessica>.

@@

One of the key determinants of efficient public services provision is a robust accountability system [@besley2003incentives; @cameron2004public; @ferejohn1986incumbent; @o1998horizontal]. In its standard definition, accountability is understood as the process of holding authorities responsible for their actions [@finer1941administrative; @mulgan2000accountability; @o1990bureaucratic]. Past studies show that accountability has a positive impact on governance, as it ensures that politicians act on behalf of voters [@freire2010ngp; @moncrieffe1998reconceptualizing], reduces the opportunities for rent-seeking and corruption [@deininger2005does; @wenar2006accountability] and improves the efficiency of the public sector [@adsera2003you; @bjorkman2009power]. Recent research also suggests accountability leads to higher economic growth because it limits state discretion in the economy and increases long-term investments in human capital [@benhabib2010economic; @suebvises2018social; @ponzetto2018social].

However, accountability systems take many forms. One promising model is that of _bottom-up monitoring_, in which citizens receive information about the shortcomings of a given project so they can evaluate and pressure underperforming public officials [@kosack2014does; @molina2016community; @raffler2018weakness]. Proponents argue bottom-up accountability is effective because: 1) constituents have first-hand information about the outcomes of local policies; 2) citizens have incentives to attack corruption that directly affects them; 3) policy-makers are sensitive to social punishment from their own communities [@serra2011combining, 570]. In a nutshell, bottom-up accountability offers a potential solution to the principal-agent dilemma in public service by aligning the interests of state officials with those of the constituency they serve [@barro1973control; @raffler2018weakness, 2].

Here we assess the impact of _Tá de Pé_ (TDP), a mobile phone application designed to lower the costs of evaluating public works and punishing political representatives in Brazil. Developed by Transparência Brasil[^transparencia], TDP allows Brazilian citizens to know the location of public school construction sites, check their completion status and anonymously request information from competent authorities. TDP users can also take pictures of the construction sites and submit them to volunteering engineers for examination. If the engineers classify the construction as delayed, TDP prompts users to send a message to the mayor's office asking for completion estimates and explanations about the construction status. TDP has been online since April 2017 in 1030 Brazilian municipalities and was the winner of the 2016 Google Social Impact grant with more than 200,000 popular votes[^google].

@@

[^google]: About 1,000 Brazilian charities participated in the 2016 Google Social Impact Challenge. An independent committee selected 10 organisations as finalists, and Transparência Brasil won the challenge with about 200,000 popular votes. To know more about the contest, please visit <https://impactchallenge.withgoogle.com/brazil2016> (access: July 2019).

We use the TDP app to conduct two experimental interventions. We test whether the app has a positive impact on three outcomes of interest: school completion rates, school construction status and complaints to competent authorities. We find that our experiments have delivered mixed results. Two of the findings are positive and in line with our theoretical predictions. From August 2017 to February 2019, TDP increased the likelihood of school completion by 6.8 percent (SE = X.X, t-test = XXX). Additionally, we find that pressure over the bureaucracy increased the investments in the schools by 157.5 percent (SE = X.XX, t-test = XXX). In contrast, our results show the app has a null impact on school constructions statistics. Brazilian politicians did not respond to the citizen's requests placed via the application.

Our findings raise questions about the ability of citizens to hold representatives accountable using bottom-up strategies. Although

**Nota**: _Reescrever os dois últimos parágrafos e colocar mais explicações sobre possíveis mecanismos, etc_

@@

Transparência Brasil partnered with the Brazilian branch of _Engineers without Borders_ (EWB), a non-governmental organisation[^ewb], to provide technical evaluations of school completion rates. The EWB bases their assessment on photos and GPS location sent through the app. The engineers' reports are later uploaded to the database and stored on the users' mobile devices so citizens can track the progress of the constructions.

@@

The TDP has also received feedback from Brazilian computer programmers and policy analists. In 2017 and 2018, Transparência Brasil announced two team programming competitions, called 'Tá de Pé Hackathons', where contributors could fix code bugs and suggest new functionalities to the TDP project. One of these innovations consists of a Twitter bot (<https://twitter.com/tadepeapp>) which posts a message on the social network each time a user submits a new picture for evaluation. This allows interested parties to track the progress of construction sites as well as add new public works to the TDP database.

# Data and Methods
\label{sec:methods}

We assessed the impact of the TDP app on
